Hybrid Threads \\ for the \\ Parrot Virtual Machine

### 2012-06-28 Paper.tex
Absolute final fixes

\tikzstyle{block} = [rectangle, draw, text width=12.3em, text centered, rounded corners, minimum height=2em]
\tikzstyle{decision} = [diamond, draw, text width=4.5em, text badly centered, node distance=2cm, inner sep=0pt]
\tikzstyle{line} = [draw, -latex']

@@

\chapter{Introduction and Motivation}

@@

\chapter{Concurrency in other Programming Platforms}

@@

\section{Java}

@@

Even though many threads append to the same array, they do not have to take any locks. The GIL implicitly protects the array by allowing only one Python instruction to be run at any time.

@@

The current design of Parrot uses four sets of registers where each one has an unlimited number of registers. The four sets are correspondingly to Parrot's types:
%
\begin{itemize}
\item integer,
\item floating point,
\item string and
\item polymorphic container (PMC).
\end{itemize}

@@

\section{Garbage Collector (GC)}

@@

\chapter{Design of Hybrid Threads}
\label{cha:design}

@@

\section{Critical Sections}

@@

\section{Garbage Collection (GC)}

@@

\chapter{Implementation of Hybrid Threads}

@@

The process of executing and preempting tasks is pictured in figure~\ref{controlflow} and described in detail in the rest of this section.

\begin{figure}
\centering
\begin{tikzpicture}[node distance = 1cm, auto]
\node[block](outerrl){Parrot\_cx\_outer\_runloop};
\node[block, below of=outerrl](nexttask){Parrot\_cx\_next\_task};
\node[block, below of=nexttask](runloop){runloop};
\node[decision, below of=runloop] (branch) {branch?};
\node[block, below of=branch, node distance=2cm](checksched){Parrot\_cx\_check\_scheduler};
\node[block, below of=checksched](runsched){Parrot\_cx\_run\_scheduler};
\node[block, right of=runsched, node distance=13.8em](checkquant){Parot\_cx\_check\_quantum};
\node[decision, below of=runsched] (taskdone) {Quantum done?};
\node[block, below of=taskdone, node distance=2cm](preempt){Parrot\_cx\_preempt\_task};
\node[block, right of=preempt, node distance=13.8em](stop){Parrot\_cx\_stop\_task};
\path[line](outerrl) -- (nexttask);
\path[line](nexttask) -- (runloop);
\path[line](runloop) -- (branch);
\path[line](branch.south) to node[right] {yes} (checksched.north);
\path[line](branch.west) to [out=180,in=220] node[right] {no} (runloop.west);
\path[line](checksched) -- (runsched);
\path[line](runsched) -- (checkquant);
\path[line](runsched) -- (taskdone);
\path[line](taskdone.south) to node[right] {yes} (preempt.north);
\path[line](taskdone.west) to [out=180,in=220] node[right] {no} (runloop.west);
\path[line](preempt) -- (stop);
\draw[->] (preempt.west) to [out=120,in=240] (outerrl.west);
\end{tikzpicture}
\caption{Control flow of task execution and preemption.}
\label{fig:controlflow}
\end{figure}

@@

\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end, respectively the beginning of the task\_queue. \textit{shift\_pmc} is used to fetch and remove a task from the tip of the \textit{task\_queue}. \textit{get\_integer} returns the number of tasks in the \textit{task\_queue}. These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread safe. Thus, the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.

@@

The Task PMC contains the following attributes:
%
\begin{itemize}
\item \textit{UINTVAL id}: Unique identifier for the task.
\item \textit{FLOATVAL birthtime}: the creation time stamp of the task.
\item \textit{Parrot\_Interp interp}: the interpreter which created the task.
\item \textit{PMC *code}: the code to run.
\item \textit{PMC *data}: additional data for the task given as parameter to \textit{code}.
\item \textit{INTVAL killed}: flag marking killed tasks.
\item \textit{PMC *waiters}: tasks waiting on this one.
\item \textit{PMC *shared}: list of variables shared with this task.
\item \textit{PMC *partner}: copy of this task on the other side of a GC barrier, meaning in another thread.
\end{itemize}

@@

\section{Alarms and Timers}

@@

\subsection{Garbage Collection}

@@

Before the implementation of threading support, having multiple interpreter objects in the system was used to implement things like a safe compartment where code can be run in a restricted environment. For such cases it was useful to share the GC between interpreters. This special handling had to be removed.

@@

\chapter{Tests and Benchmarks}

@@

\begin{table}
\caption{Runtime comparison for Mandelbrot set calculation}
\label{benchmark:mandel}
\begin{tabular}{|l||c||c|c|c|c|} \hline
& \emph{singlethreaded} & \emph{1 thread} & \emph{2 threads} & \emph{4 threads} & \emph{8 threads} \\
\hline
1. run  & 89.931s & 89.978s & 45.813s & 24.028s & 17.445s \\
\hline
2. run  & 89.707s & 89.871s & 45.906s & 24.048s & 17.695s \\
\hline
3. run  & 90.318s & 89.839s & 45.951s & 24.049s & 17.573s \\
\hline
average & 89.985s & 89.896s & 45.890s & 24.042s & 17.571s  \\
\hline
speedup & 1       &  1      &  1.959  &  3.739  &  5.116  \\
\hline
\end{tabular}
\end{table}

@@

\chapter{Conclusion, Further Work and Experiences}

@@

\section{Further Work}

@@

\appendix
\chapter{moretasks.pir}
\label{app:moretasks}

@@

\chapter{matrix\_part.winxed}
\label{app:matrix_part}

@@

\chapter{chameneos.pir}
\label{app:chameneos_part}

>>>>

\tikzstyle{endpoint} = [circle, draw, text width=1.8em, text centered, rounded corners, minimum height=1.8em]
\tikzstyle{block} = [rectangle, draw, text width=12.3em, text centered, rounded corners, minimum height=2em]
\tikzstyle{decision} = [diamond, draw, text width=4.5em, text badly centered, node distance=1.9cm, inner sep=0pt]
\tikzstyle{line} = [draw, -latex']

@@

\chapter{Introduction and motivation}

@@

\chapter{Concurrency in other programming platforms}

@@

\section{Java}
\label{sec:java}

@@

Even though many threads append to the same array, they do not have to take any locks. The GIL implicitly protects the array by allowing only one Python instruction to be run at any time. This is a contrast to the Java version in section~\ref{sec:java} where the user has to prevent concurrency issues either by using a thread-safe data structure or through manual locking.

@@

The current design of Parrot uses four sets of registers where each one has an unlimited number of registers. The four sets are correspondingly to Parrot's types:
%
\begin{itemize}
\item integer,
\item floating point,
\item string and
\item PolyMorphic Container (PMC).
\end{itemize}

@@

\section{Garbage collector (GC)}

@@

\chapter{Design of hybrid threads}
\label{cha:design}

@@

\section{Critical sections}

@@

\section{Garbage collection (GC)}

@@

\chapter{Implementation of hybrid threads}

@@

The process of executing and preempting tasks is pictured in figure~\ref{fig:controlflow} and described in detail in the rest of this section.

\begin{figure}
\centering
\begin{tikzpicture}[node distance = 1cm, auto]
\node[endpoint](start){start};
\node[block, below of=start, node distance=1.1cm](outerrl){Parrot\_cx\_outer\_runloop};
\node[decision, below of=outerrl] (moretasks) {tasks available?};
\node[block, below of=moretasks, node distance=2cm](nexttask){Parrot\_cx\_next\_task};
\node[endpoint, right of=moretasks, node distance = 13.8em](end){end};
\node[block, below of=nexttask](runloop){runloop};
\node[decision, below of=runloop, node distance=1.7cm] (branch) {branch?};
\node[block, below of=branch, node distance=2cm](checksched){Parrot\_cx\_check\_scheduler};
\node[block, below of=checksched](runsched){Parrot\_cx\_run\_scheduler};
\node[block, right of=runsched, node distance=13.8em](checkquant){Parot\_cx\_check\_quantum};
\node[decision, below of=runsched] (taskdone) {Quantum done?};
\node[block, below of=taskdone, node distance=2cm](preempt){Parrot\_cx\_preempt\_task};
\node[block, right of=preempt, node distance=13.8em](stop){Parrot\_cx\_stop\_task};
\path[line](start) -- (outerrl);
\path[line](outerrl) -- (moretasks);
\path[line](moretasks.south) to node[right] {yes} (nexttask.north);
\path[line](moretasks.east) to  node[below] {no} (end.west);
\path[line](nexttask) -- (runloop);
\path[line](runloop) -- (branch);
\path[line](branch.south) to node[right] {yes} (checksched.north);
\path[line](branch.west) to [out=180,in=230] node[right] {no} (runloop.west);
\path[line](checksched) -- (runsched);
\path[line](runsched) -- (checkquant);
\path[line](checkquant)[yshift=-0.3cm] -- (runsched)[yshift=-0.3cm];
\path[line](runsched) -- (taskdone);
\path[line](taskdone.south) to node[right] {yes} (preempt.north);
\path[line](taskdone.west) to [out=150,in=230] node[right] {no} (runloop.west);
\path[line](preempt) -- (stop);
\path[line](stop)[yshift=-0.3cm] -- (preempt)[yshift=-0.3cm];
\draw[->] (preempt.west) to [out=110,in=250] (outerrl.west);
\end{tikzpicture}
\caption{Control flow of task execution and preemption.}
\label{fig:controlflow}
\end{figure}

@@

\textit{push\_pmc} and \textit{unshift\_pmc} are used to add a task at the end, respectively the beginning of the task\_queue. \textit{shift\_pmc} is used to fetch and remove a task from the tip of the \textit{task\_queue}. \textit{get\_integer} returns the number of tasks in the \textit{task\_queue}. These four methods use the \textit{task\_queue\_lock} to make accessing the \textit{task\_queue} thread-safe. Thus, the Scheduler PMC acts as a container while using \textit{task\_queue} to actually store the data.

@@

The Task PMC contains the following attributes:
%
\begin{itemize}
\item \textit{UINTVAL id}: Unique identifier for the task.
\item \textit{FLOATVAL birthtime}: the creation time stamp of the task.
\item \textit{Parrot\_Interp interp}: the interpreter which created the task.
\item \textit{PMC *code}: the code to run.
\item \textit{PMC *data}: additional data for the task given as parameter to \textit{code}.
\item \textit{INTVAL killed}: flag marking killed tasks.
\item \textit{PMC *waiters}: tasks waiting on this one.
\item \textit{PMC *shared}: list of variables shared with this task.
\item \textit{PMC *partner}: copy of this task in another thread.
\end{itemize}

@@

\section{Alarms and timers}

@@

\subsection{Garbage collection}

@@

@@

\chapter{Tests and benchmarks}

@@

\begin{table}
\caption{Runtime comparison for Mandelbrot set calculation}
\label{benchmark:mandel}
\begin{tabular}{|l||c||c|c|c|c|} \hline
& \emph{singlethreaded} & \emph{1 thread} & \emph{2 threads} & \emph{4 threads} & \emph{8 threads} \\
\hline
1. run  & 89.931 s & 89.978 s & 45.813 s & 24.028 s & 17.445 s \\
\hline
2. run  & 89.707 s & 89.871 s & 45.906 s & 24.048 s & 17.695 s \\
\hline
3. run  & 90.318 s & 89.839 s & 45.951 s & 24.049 s & 17.573 s \\
\hline
average & 89.985 s & 89.896 s & 45.890 s & 24.042 s & 17.571 s  \\
\hline
speedup &  1.000  &  1.001  &  1.959  &  3.739  &  5.116  \\
\hline
\end{tabular}
\end{table}

@@

\chapter{Conclusion, further work and experiences}

@@

\section{Further work}

@@

\appendix
\chapter{\textit{moretasks.pir}}
\label{app:moretasks}

@@

\chapter{\textit{matrix\_part.winxed}}
\label{app:matrix_part}

@@

\chapter{\textit{chameneos.pir}}
\label{app:chameneos_part}

### 2012-06-28 Praktikum.tex
Absolute final fixes

Atikon EDV \& Marketing GmbH entwickelt Webseiten hauptsächlich für Steuerberater und Ärzte, sowie Software für Steuerberater. Als Leiter der Webprogrammierung führe ich ein Team von 5 Programmierern. Wir entwickeln Software für den internen Gebrauch sowie für externe Kunden, z.B. das Content Management System, die Projektverwaltung, Online-Steuerberechnung und Individualentwicklung wie z.B. ein CRM System für einen unserer Kunden.

@@

Meine Aufgaben waren dabei neben der Projektleitung:
%
\begin{itemize}
\item{Entwurf der Architektur des neuen Systems}
\item{Implementierung der Grundstruktur}
\item{Implementierung des Exports aus dem alten CMS}
\item{Implementierung der Unterstützung für die Templatesprache des alten CMS}
\item{Erweiterung des Inline::Python Perlmoduls um die von uns benötigten Features}
\item{Implementierung von automatischen Kompatibilitätstests mittels Screenshotvergleich der Webseiten mit allen Unterseiten.}
\end{itemize}

@@

Die Objekte einer Website werden in einem Baum verwaltet. Um Abfragen auf Teilbäume effizienter zu machen, werden in der Tabelle \textit{zms\_object\_tree} Informationen über die Baumstruktur gespeichert. Sie erlaubt effiziente Abfragen auf komplette Unterbäume mit optionaler Begrenzung der Suchtiefe. Die Informationen werden von Triggern in der Datenbank automatisch bei Änderungen an den Daten aktualisiert.

@@

Die von den Inhaltsobjekten unterstützten Attribute sind als von \textit{ZMS::Attribute} abgeleitete Klassen implementiert. Sie enthalten Informationen über den möglicherweise benutzten Datentyp eines Datenbankfeldes, Code zum Rendern des Eingabeinterfaces und für die Eingabevalidierung.

\textit{Catalyst} gibt eine Grundstruktur für den Aufbau der Anwendung mit Unterteilung in Models, Views und Controllern vor. \textit{ZMS::Model::DB} implementiert alle low level Datenbankzugriffe und generiert daher auch alle verwendeten SQL Statements. Um die Ausgabe der Website kümmert sich \textit{ZMS::Controller::Content}. Die Verwaltung der Inhalte wird von \textit{ZMS::Controller::Content::Management} gestellt. Die Verwaltung der Metadaten wie Typen, Sprachen und Benutzer finden sich im \textit{ZMS::Controller::System} Namensraum.

>>>>

Atikon EDV \& Marketing GmbH entwickelt Webseiten hauptsächlich für Steuerberater und Ärzte, sowie Software für Steuerberater. Als Leiter der Webprogrammierung führe ich ein Team von 5 Programmierern. Wir entwickeln Software für den internen Gebrauch sowie für externe Kunden, z.B. das Content Management System, die Projektverwaltung, Online-Steuerberech\-nung und Individualentwicklung wie z.B. ein CRM System für einen unserer Kunden.

@@

Meine Aufgaben waren dabei neben der Projektleitung:
%
\begin{itemize}
\item{Entwurf der Architektur des neuen Systems}
\item{Implementierung der Grundstruktur}
\item{Implementierung des Exports aus dem alten CMS}
\item{Implementierung der Unterstützung für die Templatesprache des alten CMS}
\item{Erweiterung des Inline::Python Perlmoduls um die von uns benötigten Features}
\item{Implementierung von automatischen Kompatibilitätstests mittels Vergleich von Screenshots der Webseiten mit allen Unterseiten.}
\end{itemize}

@@

Die Objekte einer Website werden in einem Baum verwaltet. Um Abfragen auf Teilbäume effizienter zu machen, werden in der Tabelle \textit{zms\_object\_\-tree} Informationen über die Baumstruktur gespeichert. Sie erlaubt effiziente Abfragen auf komplette Unterbäume mit optionaler Begrenzung der Suchtiefe. Die Informationen werden von Triggern in der Datenbank automatisch bei Änderungen an den Daten aktualisiert.

@@

Die von den Inhaltsobjekten unterstützten Attribute sind als von \textit{ZMS::\-Attribute} abgeleitete Klassen implementiert. Sie enthalten Informationen über den möglicherweise benutzten Datentyp eines Datenbankfeldes, Code zum Rendern des Eingabeinterfaces und für die Eingabevalidierung.

\textit{Catalyst} gibt eine Grundstruktur für den Aufbau der Anwendung mit Unterteilung in Models, Views und Controllern vor. \textit{ZMS::Model::DB} implementiert alle low level Datenbankzugriffe und generiert daher auch alle verwendeten SQL Statements. Um die Ausgabe der Website kümmert sich \textit{ZMS::Controller::Content}. Die Verwaltung der Inhalte wird von \textit{ZMS::Con\-troller::Content::Management} gestellt. Die Verwaltung der Metadaten wie Typen, Sprachen und Benutzer finden sich im \textit{ZMS::Controller::System} Namensraum.
